---
title: "Hubway Visualization Challenge"
author: "Nam Pham"
date: "December 18, 2016"
output:
  html_document: default
  pdf_document: default
vertical_layout: scroll
---

```{r library, message=FALSE,warning=FALSE}
library(knitr)
library(dplyr)
library(readr)
library(ggplot2)
library(maptools)
#library(RNeo4j)
library(stringr)
library(ggmap)
library(kableExtra)
```


```{r import, message=FALSE,warning=FALSE}
csv.folder = "C:/Users/namhpham/Documents/Personal files/R workspace"
file.stations = paste(csv.folder, "hubway_stations.csv", sep="/")
file.trips    = paste(csv.folder, "hubway_trips.csv"   , sep="/")
df.stations = read_csv(file=file.stations)
df.trips    = read_csv(file=file.trips)
```


```{r modify trip dataset, message= FALSE, warning=FALSE}
df.trips %>%
  filter(!is.na(strt_statn), 
         !is.na( end_statn),
         duration > 300) %>%           #include trip > 1 minute
  mutate(age=2013-birth_date) %>%     #add age variable at the time of dataset
  mutate(raw.date=as.Date(gsub( " .*$", "", start_date), "%m/%d/%Y")) %>% 
  #extract the date before the white space and convert to Date type
  mutate(duration=duration/60)%>% #convert duration to minutes
  mutate(zip_code=str_pad((gsub("'", '',zip_code)), 5, pad="0")) %>% 
  #remove ' in the beginning and add a leading 0 if zipcode has less than 4 letter
  select(seq_id, hubway_id, strt_statn, end_statn, 
         start_date, end_date, subsc_type,
         duration, zip_code, birth_date, gender, age, raw.date, bike_nr)%>%
         {.}->df.trips
```

```{r Modify station file}
df.trips%>%
  group_by(end_statn)%>%
  summarize(incoming_trip=length(end_statn))-> count_trips
  #count number of times that end_station appears, grouped by itself

df.stations%>%
  mutate(incoming_trip=count_trips$incoming_trip)%>%
  #add new column of incoming trips to existing station, because station_id in both tables are arranged from 3->145, making it possible to merge
  {.}->df.stations.graph
```


### Greater Boston area is adopting Hubway biking system. Since 2011, number of trips has increased over time, with heavy concentration in summer months
```{r Trips}
df.trips %>%
  mutate(month_year=format(raw.date,"%Y-%m")) %>%
  ggplot(aes(x=month_year,y=(unique(seq_id))))+
  #scale_x_date(format="%Y-%m")+
  stat_summary(fun.y=length,geom="bar",fill="dodgerblue1",color="blue")  +
  #count the length of unique trip id, which is total trips
  theme(axis.text.x=element_text(angle=45,hjust=1))+
  #rotate the x-axis by 90 degree for easier viewing
  ggtitle("Total Monthly Trips from 2011 to 2013") + 
  scale_y_continuous(labels = scales::comma) +
  theme(panel.background = element_rect(fill='white', colour='black'))+
  ylab("Number of Trips")
```       



### Hubway usage time is greater on weekend, while weekday trips are shorter
```{r Days of week, result='asis',warning=FALSE, message=FALSE, tidy=TRUE, echo=FALSE, results='markup'}
options (digits = 4)
df.trips %>%
  mutate(Days_of_Week=weekdays(raw.date)) %>% #extract weekday from raw date
  group_by(Days_of_Week)%>%
  summarize(Average_Minutes=mean(duration,na.rm=TRUE),
            'Average Trips'=n()/unclass(difftime(tail(df.trips$raw.date,n=1),df.trips$raw.date[1],units="weeks"))) %>%
  arrange(desc(Average_Minutes)) %>%
  knitr::kable(.)%>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```




### Typical users of Hubway biking system are males in their 30s
    
```{r}
df.trips %>%
  filter(!is.na(gender),!is.na(age))%>%
  ggplot(aes(age)) + 
  geom_histogram(binwidth=1, fill="orange")+
  #geom_density(col="blue")+
  ylab("Number of Users")+
  xlab("Age")+
  geom_bar(aes(fill=gender)) +
  theme(panel.background = element_rect(fill='white', colour='black'))+
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Age Distribution")
```

   



### Most weekday bikers are registered users, who presumably use Hubway frequently to commute to work. Weekend trips are more utilized by one-time, casual users
```{r}
df.trips %>%
  #mutate(duration_type=ifelse(df.trips$duration >=30, "Long Trip", "< 30 min"))%>%
  mutate(Weekday=weekdays(raw.date)) %>% #extract weekday from raw date
  #if duration is longer than 20 minutes, then "Long trip", otherwise it is short
  ggplot  (aes(x   =Weekday))+
  scale_x_discrete(limits=c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"))+
  geom_bar(aes(fill=subsc_type)) +
  ylab("Number of Trips")+
  scale_y_continuous(labels = scales::comma) +
  xlab("Days of Week") +
  theme(panel.background = element_rect(fill='white', colour='black'))+
  ggtitle("Trips Distribution")
```



### Boston and Cambridge are the hotspots for Hubway. These two cities account for the highest incoming trips, from both out-of-city and within-city  {data-height=1500}
```{r GG Map, fig.align='top', tidy=TRUE, warning = FALSE}
qmap(location = "Back Bay, MA", zoom=13, maptype= "terrain" ) + 
  geom_point(data=df.stations.graph,aes(x=lng,y=lat,col=municipal, size=incoming_trip))+
  ggtitle("Top Destinations for Hubway users")
```
Biking is especially popular along Commonwealth Ave, where protected bike lanes are installed

### Improving the System
- Add more bikes to Somerville and Brookline areas
- Install more bike racks in Financial District/South Station (Boston) and Harvard/MIT (Cambridge),  the stations with top incoming traffics
- Expand the Hubway system along bike trails and tourist attractions, such as Northern Strand Community Trail (Bike to the Sea Trail)
